\documentclass[11pt,a4paper]{article}

% Isabelle document support
\usepackage{isabelle,isabellesym}

% Text + monospace quality
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{inconsolata}

% Hyperlinks
\usepackage{hyperref}

% Safe code listings for inline examples in the prose
\usepackage{listings}
\lstset{
  basicstyle=\ttfamily\small,
  columns=fullflexible,
  keepspaces=true,
  showstringspaces=false,
  upquote=true,
  frame=single,
  breaklines=true
}

\title{Advaita Vedānta: A Formalization in Isabelle/HOL}
\author{Matthew Scherf}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
This entry presents a compact Isabelle/HOL formalization of a minimalist core of
Advaita Vedānta. The development isolates neutral higher\=/order predicates for
awareness, appearance, and inseparability, together with a distinguished ground
corresponding to Brahman. Within classical HOL we axiomatize non\=/duality
(\emph{Atman = Brahman}) via a small collection of conservative assumptions and
derive representative consequences: uniqueness of the ground, dependence of
appearances on awareness, non\=/reification of awareness, and stability
properties. Proofs are short and deterministic, suitable for AFP batch builds,
and the theory is designed to be reusable alongside related non\=/dual
formalizations.
\end{abstract}

\tableofcontents
\bigskip

\section{Introduction}

Advaita Vedānta asserts that the ultimate reality (Brahman) and the innermost
self (Ātman) are not two. This entry extracts a modest, logically explicit core
of that view that can be checked mechanically in Isabelle/HOL. The purpose is
expository and foundational: specify assumptions with mathematical clarity,
study immediate consequences, and provide a small, auditable scaffold for
comparison with neighboring non\=/dual systems.

\section{Entry Overview}

The session consists of a single theory \texttt{Advaita\_Vedanta} importing
\texttt{Main}. At a high level it introduces:
\begin{itemize}
  \item a background type of entities;
  \item a distinguished constant intended as the ground (Brahman);
  \item predicates for awareness, appearance (phenomenal content), and
        inseparability;
  \item optional structure such as a symmetry/action on appearances;
  \item axioms encoding uniqueness of the ground, self\=/luminosity of
        awareness, and the non\=/duality schema.
\end{itemize}
The formalization emphasizes transparency over expressiveness and keeps proofs
short (\texttt{auto}, \texttt{blast}, \texttt{metis}).

\section{Logical Design}

\subsection{Core Symbols (Sketch)}

We work in classical HOL with a small signature. A typical abridged header:
\begin{lstlisting}
typedecl entity

consts
  Brahman     :: entity
  Awareness   :: "entity ⇒ bool"      (* Atman-as-awareness surrogate *)
  Appearance  :: "entity ⇒ bool"      (* phenomenal content / maya *)
  Inseparable :: "entity ⇒ entity ⇒ bool"
  act         :: "('g) ⇒ entity ⇒ entity"   (* optional symmetry on appearances *)
\end{lstlisting}

\subsection{Axiom Sketch}

The axioms are intentionally minimal and phenomenology-friendly:
\begin{itemize}
  \item \emph{Uniqueness of ground:} Brahman is the unique ground.
  \item \emph{Self\=/luminosity / non\=/reification:} awareness is not an appearance.
  \item \emph{Non\=/duality schema:} appearances are inseparable from the ground
        as known in awareness (a neutral surrogate for \emph{Atman = Brahman}).
  \item \emph{Dependence of appearances:} every appearance is inseparable from
        Brahman.
  \item \emph{Preservation:} optional closure/preservation of inseparability
        under symmetry/action on appearances.
\end{itemize}
All are stated as plain HOL formulas; no non\=/classical meta\=/logic is required.

\section{Illustrative Results}

\paragraph{Appearance Depends on the Ground.}
Every appearance is inseparable from Brahman:
\begin{lstlisting}
lemma appearance_inseparable_from_brahman:
  assumes "Appearance x"
  shows   "Inseparable x Brahman"
\end{lstlisting}

\paragraph{Awareness Is Not an Appearance.}
Non\=/reification of awareness:
\begin{lstlisting}
lemma awareness_not_appearance:
  assumes "Awareness a"
  shows   "¬ Appearance a"
\end{lstlisting}

\paragraph{Symmetry Preserves Non\=/Duality.}
Optional stability under action:
\begin{lstlisting}
lemma symmetry_preserves_inseparability:
  assumes "Appearance x"
  shows   "Inseparable (act g x) Brahman"
\end{lstlisting}

\paragraph{Uniqueness of the Ground.}
No second ground distinct from Brahman:
\begin{lstlisting}
lemma ground_unique:
  assumes "Inseparable Brahman y" "Inseparable y Brahman"
  shows   "y = Brahman"
\end{lstlisting}

The proofs in the entry instantiate these sketches with the specific axioms in
\texttt{Advaita\_Vedanta.thy} and close with standard methods (\texttt{blast},
\texttt{metis}, or \texttt{auto}); no interactive tools run during batch builds.

\section{Methods and Sanity Checks}

During development, model exploration (e.g.\ Nitpick) and proof search
(Sledgehammer) may be used interactively to tune axioms and craft short proofs,
but the submitted theory contains only deterministic proofs suitable for AFP.
Any such exploratory settings are documented in comments and omitted from batch
invocation.

\section{Relation to Prior Work}

This entry complements small, reusable formalizations of non\=/dual frameworks
(e.g.\ secular non\=/duality, Daoism, Dzogchen) by providing an Advaita\=/specific
core within the same HOL substrate. The shared setting enables side\=/by\=/side
comparison and potential unification via interface theories.

\section{Limitations and Future Work}

The axioms aim for faithfulness and neutrality but necessarily abstract from the
full exegetical and soteriological context. Natural extensions include:
\begin{itemize}
  \item refining the \texttt{Inseparable} relation toward epistemic or modal
        accounts of \emph{being known};
  \item pinning down algebraic laws for \texttt{act} and studying equivariance;
  \item conservative embeddings into richer libraries for cross\=/domain
        comparisons.
\end{itemize}

\section{How to Build}

From the session directory:
\begin{lstlisting}
isabelle build -c -o document=pdf Advaita_Vedanta
\end{lstlisting}
To place the PDF alongside sources:
\begin{lstlisting}
isabelle build -c -D . -o document=pdf -o document_output=public Advaita_Vedanta
\end{lstlisting}

\bigskip

% ---------------------------------------------------------------------------

\input{session}

\end{document}
